Agda Meets Accelerate

### 2012-08-01 paper/main.tex
section 3 4 5

% 16pages; IFL 2012
%
\documentclass{llncs}
%
\usepackage{amssymb}
\usepackage{cite}
% \usepackage[T1]{fontenc}
% \usepackage[utf8]{inputenc}

@@

@@

Agda \cite{Norell2008,BoveDybjerNorell2009} is a dependently typed
functional programming language. Its basis is a dependently typed
lambda calculus extended with inductive data type families, dependent
records, and parameterized modules. At the same time, Agda is also a
proof assistant for interactively constructing proofs in an
intuitionistic type theory based on the work of Per Martin-L\"of
\cite{MartinLoef1984}.

@@

Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011} is a data-parallel array language embedded into Haskell, which targets GPUs. It is a \emph{generative library}, as its data-parallel array operations are not executed directly, but instead construct abstract syntax trees (AST) representing an entire data-parallel subcomputation. These \emph{computation representations} are executed using a \verb+run+ operation that accepts such an AST (of type \verb+Acc a+), compiles it to GPU kernels, uploads it to the device, executes it, and retrieves the results.
%
\begin{verbatim}
CUDA.run :: Arrays a => Acc a -> a
\end{verbatim}
%
The type class constraint \verb+Arrays a+ restricts the result type to a single array or a tuple of arrays.

@@

As a simple example of an Accelerate program, consider a function implementing a dot product:
%
\begin{verbatim}
dotp :: Vector Float -> Vector Float -> Acc (Scalar Float)
dotp xs ys = let xs' = use xs
                 ys' = use ys
             in  fold (+) 0 (zipWith (*) xs' ys')
\end{verbatim}
%
The types \verb+Vector+ and \verb+Scalar+ represent one- and zero-dimensional
arrays. Plain arrays, such as \verb+Vector Float+ are conventional Haskell arrays, using an unboxed representation to improve performance. However, when they are wrapped into the constructor \verb+Acc+, such as in \verb+Acc (Scalar Float)+, they represent arrays of the embedded language and are allocated in GPU memory, which in current high-performance GPUs is physically separate from CPU memory.

@@

The type \verb+Scalar+ and \verb+Vector+ are merely type synonyms instantiating a shape-parameterised array type to the special case of zero and one dimensional arrays:
%
\begin{verbatim}
type Scalar e = Array DIM0 e
type Vector e = Array DIM1 e
\end{verbatim}
%
So, the general type for \verb+use+ is
%
\begin{verbatim}
use :: Elt e => Array sh e -> Acc (Array sh e)
\end{verbatim}
%
where the class \verb+Elt+ characterises all types that may be held in Accelerate arrays. These are currently primitive types and tuples.

Common dimensions, such as \verb+DIM0+, \verb+DIM1+, and so on, are predefined, but to enable shape polymorphic computations, along the lines pioneered in the Haskell array library Repa~\cite{keller-etal:repa}, shapes are inductively defined using type-level snoc lists formed from
%
\begin{verbatim}
data Z       = Z
data sh :. i = sh :. i
\end{verbatim}
%
Hence, we have got
%
\begin{verbatim}
type DIM0 = Z
type DIM1 = DIM0 :. Int
<and so on>
\end{verbatim}

@@

In this section, we investigate the potential uses of dependent typing
in a language like Accelerate.

@@

Accelerate's API features expressive type constraints that describe
the shape of the array arguments and results. These constraints ensure
that no shape mismatches occur (e.g., a one-dimensional array is
considered two-dimensional). However, they do not ensure at compile
time that the sizes of the dimensions match up.

As an example, consider the function \texttt{reshape}. 
It takes a target shape \texttt{sh} and an array of source shape
\texttt{sh'} and changes the layout of that array to \texttt{sh}. 
\begin{verbatim}
reshape :: Exp sh -> Acc (Array sh' e) -> Acc (Array sh e)
\end{verbatim}
For this reshaping to work correctly, the underlying
number of elements must remain the same. For example, while it makes sense
to reshape a two-dimensional $3\times 4$-array  to a vector of size
$12$ or to a three-dimensional $3\times2\times2$-array, whereas an attempt to
reshape the same array to a $2\times5$-array should be rejected at
compile time.

In Agda, we define the shape type as follows.
\begin{verbatim}
data Shape : Set where
  Z     : Shape
  _:<_> : Shape -> Nat -> Shape
\end{verbatim}
The \texttt{size} function computes the number of elements stored in a
shape.
\begin{verbatim}
size : Shape -> Nat
size Z = 1
size (sh :< n >) = size sh * n
\end{verbatim}
Now we can state an accurate type for \texttt{reshape} in Agda, which
involves an extra argument with a proof that the source and
target shapes have the same size.\footnote{For the moment, the type
  \texttt{Element} could be read as \texttt{Set}. Actually,
  \texttt{Element} classifies the types of values that can be stored
  in Accelerate arrays. In Haskell, this restriction is imposed with a
  type class.}
\begin{verbatim}
reshape : {sh : Shape} {E : Element}
       -> (sh' : Shape) -> Array sh E -> (size sh = size sh')
       -> Array sh' E
\end{verbatim}
There is a subtle difference to the original signature. In
Accelerate, the first argument is an \emph{expression} that produces a
value of type \texttt{sh} at run time, whereas the Agda
\texttt{reshape} requires a \texttt{Shape} as its first argument.

@@

Some parallel reduction operations require their base operation to be
associative to return a predictable result. Here are two examples from
Accelerate. 
\begin{verbatim}
fold  :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) -> Exp a ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
fold1 :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
\end{verbatim}
In both cases, the text of the documentation says that ``the first
argument needs to be associative'' and the \texttt{fold1}
documentation ``requires the reduced array to be non-empty''.
The second requirement can be enforced by asking for a suitable proof
object on each call of \texttt{fold1}:
\begin{verbatim}
fold1 : ... -> Array (sh :< n >) E -> (p : size sh * n > 0)
            -> Array sh E
\end{verbatim}
The first requirement can be rephrased to saying that the first two
parameters of \texttt{fold} together form a monoid, which requires an
associative operation with a unit element. The concept of a monoid
can be formalized in Agda, which has indeed been done in the standard
library. Unfortunately, the formalization from the library cannot be
used because Accelerate deals with ASTs, not with values. So, a
formalization is required that states that the meaning of an
AST-encoded function is associative and the meaning of another
AST-encoded constant is its unit element. Given that Accelerate
encodes AST construction using higher-order abstract syntax, such a
formalization is not straightforward. Moreover, even given expressions
with a fixed meaning, there is no general shape for associative
functions, so that proofs can only be done for special cases.

\subsection{Embedding of Haskell Types}
\label{sec:embedding-constants}

Accelerate supports a wide range of numeric types as base types for
array computations. Almost all of these types have no suitable
counterpart in Agda, which only supports natural numbers in unary
encoding. For that reason, our embedding keeps these types 
abstract in Agda. To specify functions that are polymorphic in such a
Haskell type or depend on it in some way, we have reified these types
in an Agda type \texttt{Element}.
\begin{verbatim}
data Element : Set where
  Bool   : Element
  Int    : Element
  Float  : Element
  Double : Element
  Pair   : Element -> Element -> Element
  ...
\end{verbatim}
Corresponding to Haskell type classes that are used in Accelerate, our
embedding supplies predicates that characterize subsets. For example,
the set of numeric type is defined by a predicate \texttt{IsNumeric}.
\begin{alltt}
IsNumeric : Element -> Set
IsNumeric Int = \(\top\)
IsNumeric Float = \(\top\)
IsNumeric Double = \(\top\)
IsNumeric _ = \(\bot\)
\end{alltt}

@@

One might contemplate that an existential type like
\begin{verbatim}
exists (\ m : Nat -> m <= n -> Vector m E)
\end{verbatim}
could be employed. However, it is not possible to build such an
existential package because the evidence \texttt{m} is not available
when the package is constructed.

@@

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End:

>>>>

% 16pages; IFL 2012
%
\documentclass{llncs}
%
\usepackage{amssymb}
\usepackage{cite}
\usepackage{color}
% \usepackage[T1]{fontenc}

@@

% UNICODE in Agda sources: more trouble than it is worth right now!

% \usepackage{bbm}
% \usepackage[greek,english]{babel}
% \usepackage{ucs}
% \usepackage[utf8x]{inputenc}
% %\usepackage{autofe}
% \DeclareUnicodeCharacter{8704}{\ensuremath{\forall}}
% % \DeclareUnicodeCharacter{8594}{\ensuremath{\mathbb{N}}}
% \DeclareUnicodeCharacter{8594}{\ensuremath{\to}}

\usepackage{fancyvrb}

\DefineVerbatimEnvironment{code}{Verbatim}{} % Add fancy options here if you like.
\DefineVerbatimEnvironment{hcode}{Verbatim}{formatcom=\color{blue},frame=leftline} % Add fancy options here if you like.

@@

Agda \cite{Norell2008,BoveDybjerNorell2009} is a dependently-typed
functional programming language. Its basis is a dependently-typed
lambda calculus extended with inductive data type families, dependent
records, and parameterized modules. At the same time, Agda is also a
proof assistant for interactively constructing proofs in an
intuitionistic type theory based on the work of Per Martin-L\"of
\cite{MartinLoef1984}.

@@

Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011} is a
data-parallel array language embedded into Haskell, which targets
GPUs. It is a \emph{generative library}, as its data-parallel array
operations are not executed directly, but instead construct abstract
syntax trees (AST) representing an entire data-parallel
subcomputation. These \emph{computation representations} are executed
using a \verb+run+ operation that accepts such an AST (of type
\verb+Acc a+), compiles it to GPU kernels, uploads it to the device,
executes it, and retrieves the results.\footnote{To distinguish
  Haskell code from Agda code, we display the former in blue and with
  a vertical bar on the left side.}
%
\begin{hcode}
CUDA.run :: Arrays a => Acc a -> a
\end{hcode}
%
The type class constraint \verb+Arrays a+ restricts the result type to a single array or a tuple of arrays.

@@

As a simple example of an Accelerate program, consider a function implementing a dot product:
%
\begin{hcode}
dotp :: Vector Float -> Vector Float -> Acc (Scalar Float)
dotp xs ys = let xs' = use xs
                 ys' = use ys
             in  fold (+) 0 (zipWith (*) xs' ys')
\end{hcode}
%
The types \verb+Vector+ and \verb+Scalar+ represent one- and zero-dimensional
arrays. Plain arrays, such as \verb+Vector Float+ are conventional Haskell arrays, using an unboxed representation to improve performance. However, when they are wrapped into the constructor \verb+Acc+, such as in \verb+Acc (Scalar Float)+, they represent arrays of the embedded language and are allocated in GPU memory, which in current high-performance GPUs is physically separate from CPU memory.

@@

The type \verb+Scalar+ and \verb+Vector+ are merely type synonyms instantiating a shape-parameterised array type to the special case of zero and one dimensional arrays:
%
\begin{hcode}
type Scalar e = Array DIM0 e
type Vector e = Array DIM1 e
\end{hcode}
%
So, the general type for \verb+use+ is
%
\begin{hcode}
use :: Elt e => Array sh e -> Acc (Array sh e)
\end{hcode}
%
where the class \verb+Elt+ characterises all types that may be held in Accelerate arrays. These are currently primitive types and tuples.

Common dimensions, such as \verb+DIM0+, \verb+DIM1+, and so on, are predefined, but to enable shape polymorphic computations, along the lines pioneered in the Haskell array library Repa~\cite{keller-etal:repa}, shapes are inductively defined using type-level snoc lists formed from
%
\begin{hcode}
data Z       = Z
data sh :. i = sh :. i
\end{hcode}
%
Hence, we have got
%
\begin{hcode}
type DIM0 = Z
type DIM1 = DIM0 :. Int
-- <and so on>
\end{hcode}

@@

In this section, we investigate the potential uses of dependent typing
in a language like Accelerate. First, we review some basics of the
embedding.

\subsection{Embedding of Haskell Types}
\label{sec:embedding-types}

Accelerate supports a wide range of numeric types as base types for
array computations. Almost all of these types have no suitable
counterpart in Agda, which only supports natural numbers in unary
encoding. For that reason, our embedding keeps the Haskell types
abstract in Agda. To specify type signatures and in particular
functions that are polymorphic in such a Haskell type or depend on it
in some way, we have reified these types in an Agda type
\texttt{Element}.
\begin{verbatim}
data Element : Set where
  Bool   : Element
  Int    : Element
  Float  : Element
  Double : Element
  Pair   : Element -> Element -> Element
  -- and so on
\end{verbatim}
Corresponding to Haskell type classes that are used in Accelerate, our
embedding supplies predicates that characterize subsets. For example,
the set of numeric types is defined by a predicate \texttt{IsNumeric}.
\begin{alltt}
IsNumeric : Element -> Set
IsNumeric Int = \(\top\)
IsNumeric Float = \(\top\)
IsNumeric Double = \(\top\)
IsNumeric _ = \(\bot\)
\end{alltt}
The embedding declares further subsets all in the same style.

\subsection{Array Types}
\label{sec:array-types}

To see the Agda embedding in action, we translate the
dot product example from Section~\ref{sec:accelerate} to Agda.
\begin{code}
dotp : {E : Element} {{p : IsNumeric E}} {n : Nat}
     -> PreVector n E -> PreVector n E -> Scalar E
dotp{E} xs ys = 
  let xs' = use xs
      ys' = use ys
  in
  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
\end{code}
Unlike the Accelerate code, this function is polymorphic with respect
to the array element type, provided it is numeric. It furthermore
takes the length $n$ as a parameter thus ensuring that the two input
vectors have the same size. The \texttt{PreVector} type of the
arguments corresponds to the plain \texttt{Vector} type in Accelerate,
whereas the result type \texttt{Scalar E} corresponds to \texttt{Acc
  (Scalar E)}---a piece of abstract syntax.

The \texttt{use} function works as before, but its type
includes more information:
\begin{code}
use : {sh : Shape}{E : Element} -> PreArray sh E -> Array sh E
\end{code}
Like \texttt{E}, the index \texttt{sh} is now an element of an ordinary type:
\begin{verbatim}
data Shape : Set where
  Z     : Shape
  _:<_> : Shape -> Nat -> Shape
\end{verbatim}
Asking for arrays of equal shape, as in the signature of \texttt{use},
means that the arrays have to have the exact same layout.
The functions \texttt{fold}, \texttt{zipWith}, and \texttt{:::} are
discussed in the subsequent Subsections. The functions \verb|_+_| and
\verb|_*_| both have the same type:
\begin{code}
_+_ : {E : Element} {{p : IsNumeric E}} -> Exp E -> Exp E -> Exp E
\end{code}
They are restricted to arguments of numeric type and construct
abstract syntax for an addition or a multiplication by delegating to
the corresponding Accelerate functions.

@@

Accelerate's API features expressive type constraints that describe
the shape of the array arguments and results. These constraints ensure
that no shape mismatches occur (e.g., a one-dimensional array cannot be
considered two-dimensional). However, they do not ensure at compile
time that the sizes of the dimensions match up.

As an example, consider the function \texttt{reshape}. 
It takes a target shape \texttt{sh} and an array of source shape
\texttt{sh'} and changes the layout of that array to \texttt{sh}. 
\begin{hcode}
reshape :: Exp sh -> Acc (Array sh' e) -> Acc (Array sh e)
\end{hcode}
For this reshaping to work correctly, the underlying
number of elements must remain the same. For example, while it makes sense
to reshape a two-dimensional $3\times 4$-array  to a vector of size
$12$ or to a three-dimensional $3\times2\times2$-array, whereas an attempt to
reshape the same array to a $2\times5$-array should be rejected at
compile time.

As \texttt{Shape} is an ordinary data type, we can define a
\texttt{size} function that computes the number of elements stored in
an array of a certain shape.
\begin{verbatim}
size : Shape -> Nat
size Z = 1
size (sh :< n >) = size sh * n
\end{verbatim}
Now we can state an accurate type for \texttt{reshape} in Agda, which
involves an extra argument with a proof that the source and
target shapes have the same size.\footnote{For the moment, the type
  \texttt{Element} could be read as \texttt{Set}. Actually,
  \texttt{Element} classifies the types of values that can be stored
  in Accelerate arrays. In Haskell, this restriction is imposed with a
  type class.}
\begin{verbatim}
reshape : {sh : Shape} {E : Element}
       -> (sh' : Shape) -> Array sh E -> (size sh = size sh')
       -> Array sh' E
\end{verbatim}
There is a subtle difference to the original signature. In
Accelerate, the first argument is an \emph{expression} that produces a
value of type \texttt{sh} at run time, whereas the Agda
\texttt{reshape} requires a \texttt{Shape} as its first argument.

@@

Some parallel reduction operations require their base operation to be
associative to return a predictable result. Here are two examples from
Accelerate. 
\begin{hcode}
fold  :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) -> Exp a ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
fold1 :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
\end{hcode}
In both cases, the text of the documentation says that ``the first
argument needs to be associative'' and the \texttt{fold1}
documentation ``requires the reduced array to be non-empty''.
The second requirement can be enforced by asking for a suitable proof
object on each call of \texttt{fold1}:
\begin{verbatim}
fold1 : ... -> Array (sh :< n >) E -> (p : size sh * n > 0)
            -> Array sh E
\end{verbatim}
The first requirement can be rephrased to saying that the first two
parameters of \texttt{fold} together form a monoid, which requires an
associative operation with a unit element. The concept of a monoid
can be formalized in Agda, which has indeed been done in the standard
library. Unfortunately, the formalization from the library cannot be
used because Accelerate deals with ASTs, not with values. So, a
formalization is required that states that the meaning of an
AST-encoded function is associative and the meaning of another
AST-encoded constant is its unit element. Given that Accelerate
encodes AST construction using higher-order abstract syntax, such a
formalization is not straightforward. Moreover, even given expressions
with a fixed meaning, there is no general shape for associative
functions, so that proofs can only be done for special cases.

In any case, providing such information would be done by including an
additional argument that holds a suitable proof object, as in
\begin{verbatim}
fold : {E}{sh}{n} -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E)
     -> Array (sh :< n >) E -> IsMonoid f e -> Array sh E
\end{verbatim}
where
\begin{verbatim}
IsMonoid : {E} -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E) -> Set
IsMonoid f e = ( IsAssociative f , IsUnit f e)
\end{verbatim}

@@

One might contemplate that an existential type like
\begin{verbatim}
exists Nat (\ m -> m <= n -> Vector m E)
\end{verbatim}
could be employed. However, it is not possible to build such an
existential package because the evidence \texttt{m} is not available
when the package is constructed.

@@

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% coding: utf-8
%%% End:
